---
title: "GO"
author: "Yutian Li"
date: "1/26/2026"
output: html_document
---

```{r}
library(DOSE)
library(pathview)
library(clusterProfiler)
library(org.Dm.eg.db)
library("AnnotationDbi")
library(enrichplot)
library(tidyverse)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(DESeq2)
library(pheatmap)
library(tibble)
```


```{r}
#DESeq2 (muscle cluster) 
# Paths 
dir <- "/Users/yutian/Desktop/DESeq2_files/"

# read files
counts <- read.table(
  file = file.path(dir, "countData_L7T7_vs_L6C4.tsv"),
  header = TRUE,
  row.names = 1,
  sep = "\t",
  check.names = FALSE
)

coldata <- read.table(
  file = file.path(dir, "colData_L7T7_vs_L6C4.tsv"),
  header = TRUE,
  row.names = 1,
  sep = "\t",
  check.names = FALSE
)

# Ensure column order matches coldata rows
counts <- counts[, rownames(coldata)]

# Basic types
coldata$condition_group <- factor(coldata$condition_group, levels = c("control", "treated"))

# Build DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = round(as.matrix(counts)),  
  colData   = coldata,
  design    = ~ condition_group
)

# Filter low-count genes
smallestGroupSize <- 0
keep <- rowSums((counts(dds)) >= 10) >= smallestGroupSize
dds <- dds[keep, ]

# Run DESeq2
dds <- DESeq(dds)
```


```{r}
# Results: treated vs control
res <- results(dds, contrast = c("condition_group", "treated", "control"))
res_tb <- as.data.frame(res)
res_tb$gene <- rownames(res_tb)

padj.cutoff <- 0.05
lfc.cutoff  <- 1

res_sig <- res_tb[
  !is.na(res_tb$padj) &
  !is.na(res_tb$log2FoldChange) &
  res_tb$padj < padj.cutoff &
  abs(res_tb$log2FoldChange) > lfc.cutoff,
]


res_up <- res_sig[res_sig$log2FoldChange > 0, ]
res_up <- res_up[order(-res_up$log2FoldChange), ]

res_down <- res_sig[res_sig$log2FoldChange < 0, ]
res_down <- res_down[order(res_down$log2FoldChange), ]
```


```{r}
genes_to_map <- unique(c(res_up$gene, res_down$gene))

fb_map <- AnnotationDbi::select(
  org.Dm.eg.db,
  keys     = genes_to_map,
  keytype  = "FLYBASE",
  columns  = c("SYMBOL", "GENENAME")
)
fb_map <- fb_map[!duplicated(fb_map$FLYBASE), ]

annotate_res_df <- function(res_df, map_df) {
  m <- match(res_df$gene, map_df$FLYBASE)
  res_df$SYMBOL   <- map_df$SYMBOL[m]
  res_df$GENENAME <- map_df$GENENAME[m]
  res_df
}
res_up_annot   <- annotate_res_df(res_up, fb_map)
res_down_annot <- annotate_res_df(res_down, fb_map)

res_up_annot   <- res_up_annot[order(-res_up_annot$log2FoldChange), ]
res_down_annot <- res_down_annot[order(res_down_annot$log2FoldChange), ]

rownames(res_up_annot)   <- NULL
rownames(res_down_annot) <- NULL
```


```{r}
write.csv(
  res_up_annot,
  file = "DESeq2_up_all_c12removed.csv",
  row.names = FALSE
)

write.csv(
  res_down_annot,
  file = "DESeq2_down_all_c12removed.csv",
  row.names = FALSE
)
```


```{r}
#go with background fix
# 0) choose thresholds (use your own)
padj.cutoff <- 0.05
lfc.cutoff  <- 1

# 1) # universe: genes with expression (exclude baseMean==0)
bg_fbgn <- unique(res_tb$gene[!is.na(res_tb$baseMean) & res_tb$baseMean > 0])


# 2) significant set
res_sig <- res_tb[
  !is.na(res_tb$padj) &
  res_tb$padj < padj.cutoff &
  abs(res_tb$log2FoldChange) > lfc.cutoff,
]

# 3) split up / down (already FBgn)
sig_fbgn_up   <- unique(res_sig$gene[res_sig$log2FoldChange > 0])
sig_fbgn_down <- unique(res_sig$gene[res_sig$log2FoldChange < 0])

ego_up_bg <- enrichGO(
  gene     = sig_fbgn_up,
  universe = bg_fbgn,
  OrgDb    = org.Dm.eg.db,
  keyType  = "FLYBASE",
  ont      = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.1,
  readable = TRUE
)

ego_down_bg <- enrichGO(
  gene     = sig_fbgn_down,
  universe = bg_fbgn,
  OrgDb    = org.Dm.eg.db,
  keyType  = "FLYBASE",
  ont      = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.1,
  readable = TRUE
)

cluster_summary_up_bg <- data.frame(ego_up_bg)
cluster_summary_up_bg

cluster_summary_down_bg <- data.frame(ego_down_bg)
cluster_summary_down_bg


dotplot(
  ego_up_bg,
  showCategory = 40,
  font.size = 6,
) + ggtitle("DEgenes_up")


dotplot(
  ego_down_bg,
  showCategory = 20,
  font.size = 6
) + ggtitle("DEgenes_down")

```

```{r}
write.csv(
  cluster_summary_up_bg,
  file = "/Users/yutian/Desktop/GO_up_all_c12removed.csv",
  row.names = FALSE
)

write.csv(
  cluster_summary_down_bg,
  file = "/Users/yutian/Desktop/GO_down_all_c12removed.csv",
  row.names = FALSE
)
```

```{r}
pdf(
  file = "/Users/yutian/Desktop/GO_up_muscle_dotplot.pdf",
  width = 7,
  height = 6
)

dotplot(
  ego_up_bg,
  showCategory = 30,
  font.size = 6
) + ggtitle("DEgenes_up")

dev.off()

pdf(
  file = "/Users/yutian/Desktop/GO_down_muscle_dotplot.pdf",
  width = 7,
  height = 6
)

dotplot(
  ego_down_bg,
  showCategory = 30,
  font.size = 6
) + ggtitle("DEgenes_down")

dev.off()
```



```{r}
#Volcano Plot - reformatting for muscle
padj.cutoff <- 0.05
lfc.cutoff  <- 1

label_genes   <- c("FBgn0283499","FBgn0001257","FBgn0001258","FBgn0033153",
                   "FBgn0010385","FBgn0260635","FBgn0264922","FBgn0020255",
                   "FBgn0028978","FBgn0003525","FBgn0005660",
                   "FBgn0000405","FBgn0000404","FBgn0004106","FBgn0003124",
                   "FBgn0001224","FBgn0013277","FBgn0013278","FBgn0013279",
                   "FBgn0010041","FBgn0010042","FBgn0010038",
                   "FBgn0051354","FBgn0013275","FBgn0001226","FBgn0001230",
                   "FBgn0001218","FBgn0001233",
                   "FBgn0001227","FBgn0001227","FBgn0001225","FBgn0001229",
                   "FBgn0038277","FBgn0030241","FBgn0001120","FBgn0000927",
                   
                   "FBgn0038236","FBgn0004169","FBgn0003429","FBgn0265487",
                   "FBgn0051146","FBgn0023441","FBgn0003380","FBgn0051146",
                   "FBgn0086680","FBgn0020378","FBgn0264695","FBgn0086906",
                   "FBgn0034515","FBgn0005666","FBgn0287605",
                   "FBgn0051374","FBgn0265991","FBgn0035917","FBgn0000116",
                   "FBgn0027562","FBgn0038294","FBgn0030921","FBgn0032266",
                   "FBgn0001208","FBgn0036278","FBgn0001114"
                   )


label_symbols <- c("InR","ImpL2","Ldh","Gadd45",
                   "Def","Diap1","smt3","Ran",
                   "trbl","stg","Ets21C",
                   "CycB","CycA","Cdk1","polo",
                   "Hsp23","Hsp70Ba","Hsp70Bb","Hsp70Bc",
                   "GstD5","GstD6","GstD2",
                   "Hsp70Bbb","Hsp70Aa","Hsp27","Hsp68",
                   "Hsc70-3","Hsp83",
                   "Hsp67Ba","Hsp70Ab","Hsp26","Hsp67Bc",
                   "RpS5b","feo","gnu","fs(1)Ya",
                   
                   
                   "Cyp313a1","up","slo","mbl",
                   "nrm","fus","Sh","Nlg1",
                   "vvl","Sp1","Mhc","sls",
                   "CG13428","bt","lncRNA:CR46474",
                   "sals","Zasp52","Zasp66","Argk",
                   "CG10345","Mf","CG6290","CG18302",
                   "Hn","CrzR","Glt"
                   )


vol <- res_tb %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>%
  mutate(
    neglog10_padj = -log10(padj),
    status = case_when(
      padj < padj.cutoff & log2FoldChange >  lfc.cutoff ~ "Up",
      padj < padj.cutoff & log2FoldChange < -lfc.cutoff ~ "Down",
      TRUE ~ "NS"
    ),
    label = if_else(
      gene %in% label_genes,
      label_symbols[match(gene, label_genes)],
      NA_character_
    )
  )


p <- ggplot(vol, aes(x = log2FoldChange, y = neglog10_padj, color = status)) +
  geom_point(size = 2, alpha = 1) +
  scale_color_manual(values = c(NS = "grey70", Up = "red3", Down = "blue3")) +
  geom_vline(xintercept = c(-lfc.cutoff, lfc.cutoff), linetype = "dashed") +
  geom_hline(yintercept = -log10(padj.cutoff), linetype = "dashed") +
  geom_point(
    data = subset(vol, gene %in% label_genes),
    shape = 21, fill = NA, color = "black", stroke = 1.0, size = 3
  ) +
  geom_text_repel(
    aes(label = label),
    color = "black",
    size = 3,
    max.overlaps = Inf,
    na.rm = TRUE
  ) +
  coord_cartesian(ylim = c(0, 20)) +
  labs(
    x = "log2 Fold Change (treated vs control)",
    y = "-log10(adjusted p-value)",
    title = "DESeq2 Volcano"
  ) +
  theme_classic()

options(repr.plot.width = 7, repr.plot.height = 14)


print(p)
```


```{r}
#Volcano Plot - reformatting
padj.cutoff <- 0.05
lfc.cutoff  <- 1
label_genes   <- c("FBgn0283499","FBgn0001257","FBgn0001258","FBgn0033153",
                   "FBgn0010385","FBgn0260635","FBgn0264922","FBgn0020255",
                   "FBgn0028978","FBgn0003525","FBgn0005660",
                   "FBgn0000405","FBgn0000404","FBgn0004106","FBgn0003124",
                   "FBgn0001224","FBgn0013277","FBgn0013278","FBgn0013279",
                   "FBgn0010041","FBgn0010042","FBgn0010038",
                   "FBgn0051354","FBgn0013275","FBgn0001226","FBgn0001230",
                   "FBgn0001218","FBgn0001233",
                   "FBgn0001227","FBgn0001227","FBgn0001225","FBgn0001229",
                   "FBgn0038277","FBgn0030241","FBgn0001120","FBgn0000927",
                   
                   "FBgn0038236","FBgn0004169","FBgn0003429","FBgn0265487",
                   "FBgn0051146","FBgn0023441","FBgn0003380","FBgn0051146",
                   "FBgn0086680","FBgn0020378","FBgn0264695","FBgn0086906",
                   "FBgn0034515","FBgn0005666","FBgn0287605",
                   "FBgn0051374","FBgn0265991","FBgn0035917","FBgn0000116",
                   "FBgn0027562","FBgn0038294","FBgn0030921","FBgn0032266",
                   "FBgn0001208","FBgn0036278","FBgn0001114"
                   )


label_symbols <- c("InR","ImpL2","Ldh","Gadd45",
                   "Def","Diap1","smt3","Ran",
                   "trbl","stg","Ets21C",
                   "CycB","CycA","Cdk1","polo",
                   "Hsp23","Hsp70Ba","Hsp70Bb","Hsp70Bc",
                   "GstD5","GstD6","GstD2",
                   "Hsp70Bbb","Hsp70Aa","Hsp27","Hsp68",
                   "Hsc70-3","Hsp83",
                   "Hsp67Ba","Hsp70Ab","Hsp26","Hsp67Bc",
                   "RpS5b","feo","gnu","fs(1)Ya",
                   
                   
                   "Cyp313a1","up","slo","mbl",
                   "nrm","fus","Sh","Nlg1",
                   "vvl","Sp1","Mhc","sls",
                   "CG13428","bt","lncRNA:CR46474",
                   "sals","Zasp52","Zasp66","Argk",
                   "CG10345","Mf","CG6290","CG18302",
                   "Hn","CrzR","Glt"
                   )


vol <- res_tb %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>%
  mutate(
    neglog10_padj = -log10(padj),
    status = case_when(
      padj < padj.cutoff & log2FoldChange >  lfc.cutoff ~ "Up",
      padj < padj.cutoff & log2FoldChange < -lfc.cutoff ~ "Down",
      TRUE ~ "NS"
    ),
    label = if_else(
      gene %in% label_genes,
      label_symbols[match(gene, label_genes)],
      NA_character_
    )
  )


p <- ggplot(vol, aes(x = log2FoldChange, y = neglog10_padj, color = status)) +
  geom_point(size = 2, alpha = 1) +
  scale_color_manual(values = c(NS = "grey70", Up = "red3", Down = "blue3")) +
  geom_vline(xintercept = c(-lfc.cutoff, lfc.cutoff), linetype = "dashed") +
  geom_hline(yintercept = -log10(padj.cutoff), linetype = "dashed") +
  geom_point(
    data = subset(vol, gene %in% label_genes),
    shape = 21, fill = NA, color = "black", stroke = 1.0, size = 3
  ) +
  geom_text_repel(
    aes(label = label),
    color = "black",
    size = 3,
    max.overlaps = Inf,
    na.rm = TRUE
  ) +
  labs(
    x = "log2 Fold Change (treated vs control)",
    y = "-log10(adjusted p-value)",
    title = "DESeq2 Volcano"
  ) +
  theme_classic()

options(repr.plot.width = 7, repr.plot.height = 14)


print(p)
```

```{r}
#Volcano Plot - reformatting for epithelial
padj.cutoff <- 0.05
lfc.cutoff  <- 1

label_genes   <- c("FBgn0053542","FBgn0031888","FBgn0005660","FBgn0283499","FBgn0261258",
                   "FBgn0031068","FBgn0020416","FBgn0030904","FBgn0020248","FBgn0020414",
                   "FBgn0033725","FBgn0035427","FBgn0050042","FBgn0034709",
                   "FBgn0260635","FBgn0033438","FBgn0000451","FBgn0041087","FBgn0000527",
                   "FBgn0001224","FBgn0013279","FBgn0010041","FBgn0010042","FBgn0001230",
                   "FBgn0051869","FBgn0001257","FBgn0027932","FBgn0033033",
                   "FBgn0035802","FBgn0035499","FBgn0284408","FBgn0004635","FBgn0001258",
                   "FBgn0015399","FBgn0053087","FBgn0010038","FBgn0243512","FBgn0085256",
                   
                   
                   "FBgn0032299","FBgn0032008","FBgn0040211","FBgn0037697","FBgn0026268",
                   "FBgn0001114","FBgn0036992","FBgn0028939","FBgn0000658","FBgn0261244",
                   "FBgn0032666","FBgn0038422","FBgn0038236","FBgn0267033","FBgn0038930",
                   "FBgn0014073","FBgn0014073","FBgn0085450","FBgn0285917","FBgn0066101",
                   "FBgn0003651","FBgn0015600","FBgn0002945","FBgn0051300","FBgn0050377"
                   )


label_symbols <- c("upd3","Pvf2","Ets21C","InR","rgn",
                   "Alr","Idgf1","upd2","stet","Idgf3",
                   "Cpr49Ac","ckd","Cpr49Ab","Swim",
                   "Diap1","Mmp2","ect","wun2","ebony",
                   "Hsp23","Hsp70Bc","GstD5","GstD6","Hsp68",
                   "CG31869","ImpL2","Akap200","scaf",
                   "Pura","Chd64","trol","rho","Ldh",
                   "kek1","LRP1","GstD2","puc","CG34227",
                   
                   
                   
                   "CG17127","CG14277","hgo","GstZ2","antdh",
                   "Glt","Hpd","NimC2","fj","inaE",
                   "CG5758","CG14880","Cyp313a1","mamo","BomT3",
                   "Tie","Lst","Snoo","sbb","LpR1",
                   "svp","toc","nkd","CG31300","CG30377"
                   )


vol <- res_tb %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>%
  mutate(
    neglog10_padj = -log10(padj),
    status = case_when(
      padj < padj.cutoff & log2FoldChange >  lfc.cutoff ~ "Up",
      padj < padj.cutoff & log2FoldChange < -lfc.cutoff ~ "Down",
      TRUE ~ "NS"
    ),
    label = if_else(
      gene %in% label_genes,
      label_symbols[match(gene, label_genes)],
      NA_character_
    )
  )


p <- ggplot(vol, aes(x = log2FoldChange, y = neglog10_padj, color = status)) +
  geom_point(size = 2, alpha = 1) +
  scale_color_manual(values = c(NS = "grey70", Up = "red3", Down = "blue3")) +
  geom_vline(xintercept = c(-lfc.cutoff, lfc.cutoff), linetype = "dashed") +
  geom_hline(yintercept = -log10(padj.cutoff), linetype = "dashed") +
  geom_point(
    data = subset(vol, gene %in% label_genes),
    shape = 21, fill = NA, color = "black", stroke = 1.0, size = 3
  ) +
  geom_text_repel(
    aes(label = label),
    color = "black",
    size = 3,
    max.overlaps = Inf,
    na.rm = TRUE
  ) +
  coord_cartesian(ylim = c(40, 200)) +
  labs(
    x = "log2 Fold Change (treated vs control)",
    y = "-log10(adjusted p-value)",
    title = "DESeq2 Volcano"
  ) +
  theme_classic()

options(repr.plot.width = 7, repr.plot.height = 14)


print(p)
```


```{r}
ggsave(
  "volcano_40_200_epithelial.pdf",
  plot = p,
  width = 8,
  height = 7,
  units = "in"
)
```


```{r}
term_list <- c(
  "mito",
  "spindle",
  "division",
  "cytokine",
  "cell cycle"
)

keys <- str_to_lower(term_list)

go_terms_kw <- cluster_summary_up_bg %>%
  mutate(desc_lc = str_to_lower(Description)) %>%
  rowwise() %>%
  mutate(matched_kw = list(keys[str_detect(desc_lc, fixed(keys))])) %>%
  ungroup() %>%
  filter(lengths(matched_kw) > 0) %>%
  mutate(matched_kw = map_chr(matched_kw, ~ paste(.x, collapse = ", "))) %>%
  select(ID, Description, matched_kw, GeneRatio, p.adjust, Count, geneID) %>%
  arrange(p.adjust)

go_terms_kw

```

```{r}
genes_from_kw_terms <- go_terms_kw %>%
  transmute(gene = .data[["geneID"]]) %>%
  filter(!is.na(gene)) %>%
  separate_rows(gene, sep = "/") %>%
  mutate(gene = str_trim(gene)) %>%
  filter(gene != "") %>%
  distinct(gene) %>%
  pull(gene)

nrow(go_terms_kw)
length(genes_from_kw_terms)
head(genes_from_kw_terms, 80)

```
```{r}
genes_from_kw_terms_with_fc <- go_terms_kw %>%
  select(ID, Description, matched_kw, GeneRatio, p.adjust, Count, geneID) %>%
  separate_rows(geneID, sep = "/") %>%
  rename(SYMBOL = geneID) %>%
  mutate(SYMBOL = str_trim(SYMBOL)) %>%
  filter(SYMBOL != "") %>%
  left_join(
    res_up_annot %>%
      select(SYMBOL, gene, log2FoldChange, padj),
    by = "SYMBOL"
  ) %>%
  arrange(matched_kw, desc(log2FoldChange))

genes_from_kw_terms_unique_fc <- genes_from_kw_terms_with_fc %>%
  left_join(
    res_up_annot %>% select(SYMBOL, GENENAME),
    by = "SYMBOL"
  ) %>%
  group_by(SYMBOL, gene, GENENAME, log2FoldChange) %>%
  summarise(
    GO_terms = paste(unique(Description), collapse = "; "),
    matched_kw = paste(unique(matched_kw), collapse = ", "),
    n_GO_terms = n_distinct(ID),
    .groups = "drop"
  ) %>%
  arrange(desc(log2FoldChange))

head(genes_from_kw_terms_unique_fc, 80)
```
```{r}
write.csv(
  genes_from_kw_terms_unique_fc,
  "genes_from_kw_terms_unique_fc.csv",
  row.names = FALSE
)
```


```{r}
# 1) vst matrix
vsd <- vst(dds, blind = FALSE)
mat <- assay(vsd)              # rows = feature IDs (likely FBgn), cols = samples
ids <- rownames(mat)

# 2) map IDs -> SYMBOL (assume FlyBase IDs)
map <- AnnotationDbi::select(
  org.Dm.eg.db,
  keys     = ids,
  keytype  = "FLYBASE",
  columns  = c("SYMBOL")
)

# keep 1-1 mapping per ID (drop duplicates if any)
map1 <- map %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(FLYBASE, .keep_all = TRUE)

# 3) attach SYMBOL to matrix, drop unmapped
mat2 <- mat[map1$FLYBASE, , drop = FALSE]
rownames(mat2) <- map1$SYMBOL

# 4) if multiple IDs map to same SYMBOL, aggregate (mean)
mat_sym <- rowsum(mat2, group = rownames(mat2)) / as.vector(table(rownames(mat2)))

# 5) subset by your GO-derived SYMBOL list
genes_use <- intersect(genes_from_kw_terms, rownames(mat_sym))
length(genes_use)

mat_sub <- mat_sym[genes_use, , drop = FALSE]

# 6) z-score per gene and plot
mat_z <- t(scale(t(mat_sub)))
```

```{r}
pdf("GO_keywords_heatmap_muscle.pdf", width = 8, height = 12)

pheatmap(
  mat_z,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 6,
  border_color = NA
)

dev.off()
```





```{r}
term_list <- c(
  "morphogenesis"
)

keys <- str_to_lower(term_list)

go_terms_kw <- cluster_summary_up_bg %>%
  mutate(desc_lc = str_to_lower(Description)) %>%
  rowwise() %>%
  mutate(matched_kw = list(keys[str_detect(desc_lc, fixed(keys))])) %>%
  ungroup() %>%
  filter(lengths(matched_kw) > 0) %>%
  mutate(matched_kw = map_chr(matched_kw, ~ paste(.x, collapse = ", "))) %>%
  select(ID, Description, matched_kw, GeneRatio, p.adjust, Count, geneID) %>%
  arrange(p.adjust)

go_terms_kw

```
```{r}
genes_from_kw_terms_with_fc <- go_terms_kw %>%
  select(ID, Description, matched_kw, GeneRatio, p.adjust, Count, geneID) %>%
  separate_rows(geneID, sep = "/") %>%
  rename(SYMBOL = geneID) %>%
  mutate(SYMBOL = str_trim(SYMBOL)) %>%
  filter(SYMBOL != "") %>%
  left_join(
    res_up_annot %>%
      select(SYMBOL, gene, log2FoldChange, padj),
    by = "SYMBOL"
  ) %>%
  arrange(matched_kw, desc(log2FoldChange))

genes_from_kw_terms_unique_fc <- genes_from_kw_terms_with_fc %>%
  left_join(
    res_up_annot %>% select(SYMBOL, GENENAME),
    by = "SYMBOL"
  ) %>%
  group_by(SYMBOL, gene, GENENAME, log2FoldChange) %>%
  summarise(
    GO_terms = paste(unique(Description), collapse = "; "),
    matched_kw = paste(unique(matched_kw), collapse = ", "),
    n_GO_terms = n_distinct(ID),
    .groups = "drop"
  ) %>%
  arrange(desc(log2FoldChange))

head(genes_from_kw_terms_unique_fc, 80)
```

```{r}
write.csv(
  genes_from_kw_terms_unique_fc,
  "morphogenesis_genes_from_kw_terms_unique_fc.csv",
  row.names = FALSE
)
```






```{r}
#Instead of all GO term genes, keywords only
genes_use <- c("CycB","CycA","Cdk1","Ets21C",
               "InR","Ran","trbl","Gadd45",
               "polo","Cks30A", "geminin", "aurB",
               "Mhc", "sals", "sls", "Zasp52", "Zasp66")


mat_sub <- mat_sym[genes_use, , drop = FALSE]
mat_z <- t(scale(t(mat_sub)))
pdf("select_keywords_heatmap_newmuscle.pdf", width=8, height=4)

pheatmap(
  mat_z,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 10,
  fontsize_col = 10,
  border_color = NA,
  breaks = seq(-2, 2, length.out = 101)
)

dev.off()
```


```{r}
# Selected keyword genes
genes_use <- c(
  # cell cycle 
  "CycA","CycB","CycB3","CycE",
  "Cdk1","polo","aurA","aurB","Incenp",
  "mad2","pav","asp",

  # regeneration
  "Ets21C","rgn","Alr",

  # muscle differentiation
  "Mhc","sals","sls","Zasp52","Zasp66","up","mbl","bt"
)

# make sure genes exist
genes_use2 <- intersect(genes_use, rownames(mat_sym))

coldata <- as.data.frame(colData(dds))

# align metadata to matrix columns
coldata <- coldata[colnames(mat_sym), , drop = FALSE]

control_cols <- colnames(mat_sym)[coldata$condition_group == "control"]
treated_cols <- colnames(mat_sym)[coldata$condition_group == "treated"]

mean_control <- rowMeans(mat_sym[genes_use2, control_cols, drop = FALSE])
mean_treated <- rowMeans(mat_sym[genes_use2, treated_cols, drop = FALSE])

# one-column fold-change matrix
mat_fc <- matrix(
  mean_treated - mean_control,
  ncol = 1
)

rownames(mat_fc) <- genes_use2

pdf("select_keywords_one_column_fc_newmuscle.pdf", width = 3, height = 4)

pheatmap(
  mat_fc,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 10,
  fontsize_col = 10,
  border_color = NA,
  breaks = seq(-2, 2, length.out = 101)
)

dev.off()
```



```{r}
# Selected keyword genes
genes_use <- c(
  # cell cycle 
  "polo","Cks30A","CycA","asp",
  "mad2","aurB","PCNA","Cdk1","CycB","aurA"

  # muscle differentiation
  #"Mhc","sals","sls","Zasp52"
)

df_fc <- as.data.frame(res) %>%
  rownames_to_column("FLYBASE")

map <- AnnotationDbi::select(
  org.Dm.eg.db,
  keys     = df_fc$FLYBASE,
  keytype  = "FLYBASE",
  columns  = c("SYMBOL")
)

df_fc2 <- df_fc %>%
  left_join(map, by = "FLYBASE") %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(SYMBOL, .keep_all = TRUE)

genes_use2 <- intersect(genes_use, df_fc2$SYMBOL)

df_sub <- df_fc2 %>%
  filter(SYMBOL %in% genes_use2)

mat_fc <- df_sub %>%
  select(SYMBOL, log2FoldChange) %>%
  column_to_rownames("SYMBOL") %>%
  as.matrix()

mat_fc <- mat_fc[genes_use2, , drop = FALSE]


```

```{r}
pdf("select_keywords_one_column_log2fc_newmuscle.pdf", width = 3.5, height = 6)

bk <- c(
  seq(min(mat_fc, na.rm = TRUE), 0, length.out = 50),
  seq(0, max(mat_fc, na.rm = TRUE), length.out = 51)[-1]
)

cols <- c(
  colorRampPalette(c("#3b6fb6", "#ffffff"))(49),
  colorRampPalette(c("#ffffff", "#fff7bc", "#fdae6b", "#d7301f"))(50)
)

pheatmap(
  mat_fc,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 9,
  fontsize_col = 10,
  show_colnames = FALSE,
  angle_col = 0,
  border_color = NA,
  breaks = bk,
  main = "Log2FC",
  color = cols,
  cellwidth = 40,
  cellheight = 10,
  legend_breaks = c(0, 3, 6),
  legend_labels = c("0", "3", "6")

)

dev.off()
```

dev.off()
```






```{r}
# Selected keyword genes for epi
genes_use <- c(
  # morphogenesis 
  "Mmp2","kkv","scaf","rho","rn","spz3","lz","en","Wnt4","Ser",
  
  # regeneration
  "Ets21C","rgn","Alr","Idgf1"

)

df_fc <- as.data.frame(res) %>%
  rownames_to_column("FLYBASE")

map <- AnnotationDbi::select(
  org.Dm.eg.db,
  keys     = df_fc$FLYBASE,
  keytype  = "FLYBASE",
  columns  = c("SYMBOL")
)

df_fc2 <- df_fc %>%
  left_join(map, by = "FLYBASE") %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(SYMBOL, .keep_all = TRUE)

genes_use2 <- intersect(genes_use, df_fc2$SYMBOL)

df_sub <- df_fc2 %>%
  filter(SYMBOL %in% genes_use2)

mat_fc <- df_sub %>%
  select(SYMBOL, log2FoldChange) %>%
  column_to_rownames("SYMBOL") %>%
  as.matrix()

mat_fc <- mat_fc[genes_use2, , drop = FALSE]

```

```{r}
pdf("select_keywords_one_column_log2fc_newepithelial.pdf", width = 3.5, height = 6)


bk <- c(
  seq(min(mat_fc, na.rm = TRUE), 0, length.out = 50),
  seq(0, max(mat_fc, na.rm = TRUE), length.out = 51)[-1]
)

cols <- c(
  colorRampPalette(c("#3b6fb6", "#ffffff"))(49),
  colorRampPalette(c("#ffffff", "#fff7bc", "#fdae6b", "#d7301f"))(50)
)

pheatmap(
  mat_fc,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 9,
  fontsize_col = 10,
  show_colnames = FALSE,
  angle_col = 0,
  border_color = NA,
  breaks = bk,
  main = "Log2FC",
  color = cols,
  cellwidth = 40,
  cellheight = 10,
  legend_breaks = c(0, 3, 6),
  legend_labels = c("0", "3", "6")

)


dev.off()
```


```{r}

mat_fc_actual <- ifelse(
  mat_fc >= 0,
  2^mat_fc,
  -2^abs(mat_fc)
)

bk <- c(
  seq(min(mat_fc_actual, na.rm = TRUE), 0, length.out = 50),
  seq(0, max(mat_fc_actual, na.rm = TRUE), length.out = 51)[-1]
)

cols <- c(
  colorRampPalette(c("#3b6fb6", "#ffffff"))(49),
  colorRampPalette(c("#ffffff", "#fff7bc", "#fdae6b", "#d7301f"))(50)
)

pheatmap(
  mat_fc_actual,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 9,
  fontsize_col = 10,
  show_colnames = FALSE,
  angle_col = 0,
  border_color = NA,
  color = cols,
  breaks = bk,
  main = "Fold change",
  cellwidth = 40,
  cellheight = 10
)
```

```{r}
mat_fc_signed <- ifelse(
  mat_fc >= 0,
  2^mat_fc - 1,
  -(1 / (2^mat_fc) - 1)
)

bk <- c(
  seq(min(mat_fc_signed, na.rm = TRUE), 0, length.out = 50),
  seq(0, max(mat_fc_signed, na.rm = TRUE), length.out = 51)[-1]
)

cols <- colorRampPalette(c(
  "#3b6fb6",
  "#ffffff",
  "#fff7bc",
  "#fdae6b",
  "#d7301f"
))(length(bk) - 1)

pheatmap(
  mat_fc_signed,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 9,
  fontsize_col = 10,
  show_colnames = FALSE,
  border_color = NA,
  color = cols,
  breaks = bk,
  main = "Fold change",
  cellwidth = 40,
  cellheight = 10
)
```







```{r}
library(simplifyEnrichment)
library(ComplexHeatmap)

go_ids <- ego_up_bg@result$ID[!is.na(ego_up_bg@result$ID) & ego_up_bg@result$p.adjust < 0.05]
go_ids <- unique(go_ids)

sim_mat <- GO_similarity(go_ids)
```


```{r}
pdf("simplifyGO.pdf", width = 10, height = 10)
simplifyGO(sim_mat)
dev.off()

```


```{r}
ht <- Heatmap(
  sim_mat,
  name = "GO sim",
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)

draw(ht)

```



```{r}
# 1) cluster on distance = 1 - similarity
d <- as.dist(1 - sim_mat)

hc <- hclust(d, method = "average")   
k  <- 10
cl <- cutree(hc, k = k)


#  term
pick_rep <- function(mat, cl_vec) {
  reps <- character()
  for (cc in sort(unique(cl_vec))) {
    ids <- names(cl_vec)[cl_vec == cc]
    sub <- mat[ids, ids, drop = FALSE]
    score <- rowMeans(sub, na.rm = TRUE)
    reps[as.character(cc)] <- names(which.max(score))
  }
  reps
}

rep_go <- pick_rep(sim_mat, cl)
rep_go

res <- ego_up_bg@result

rep_df <- data.frame(
  cluster = as.integer(names(rep_go10)),
  GOID    = unname(rep_go10),
  stringsAsFactors = FALSE
)

m <- match(rep_df$GOID, res$ID)

rep_df$Description <- res$Description[m]
rep_df$Count       <- res$Count[m]
rep_df$padj        <- res$p.adjust[m]

rep_df <- rep_df[order(rep_df$cluster), ]
rep_df
```

```{r}
ord <- hc$order
sim_ord <- sim_mat[ord, ord, drop = FALSE]
cl_ord  <- cl10[ord]

# 标出每个 cluster 的代表 GO（右侧打标签）
at_idx <- match(rep_go10, rownames(sim_ord))
lab    <- paste0("C", names(rep_go10), ": ", rep_go10)

ht <- Heatmap(
  sim_ord,
  name = "GO sim",
  row_split = cl_ord,
  column_split = cl_ord,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = FALSE
)

# 右侧标注代表 term 的 GO ID（先标 GOID，下一步再换成 Description）
# 用 cluster -> Description 做标签
desc_map <- setNames(rep_df$Description, rep_df$cluster)

lab_desc <- paste0(
  "C", names(rep_go10), ": ",
  desc_map[as.integer(names(rep_go10))]
)

# 重新画（其余参数不变）
ht2 <- ht + rowAnnotation(
  rep = anno_mark(
    at = at_idx,
    labels = lab_desc,
    labels_gp = grid::gpar(fontsize = 8)
  )
)


draw(ht2)
```

```{r}
library(rrvgo)

go_ids <- cluster_summary_up_bg$ID

simMatrix <- calculateSimMatrix(
  go_ids,
  orgdb = "org.Dm.eg.db", 
  ont   = "BP",
  method = "Rel"
)
scores <- setNames(-log10(cluster_summary_up_bg$qvalue), cluster_summary_up_bg$ID)

reducedTerms <- reduceSimMatrix(simMatrix,
                                scores,
                                threshold=0.7,
                                orgdb="org.Dm.eg.db")


```
```{r}
treemapPlot(reducedTerms)
```

```{r}
# ---- Make treemap blocks (parents) have non-zero size ----
rt2 <- reducedTerms

# If "size" exists and some parents have size 0, assign each parent the sum of its cluster sizes
if ("size" %in% colnames(rt2)) {
  cluster_total <- tapply(rt2$size, rt2$cluster, sum, na.rm = TRUE)
  is_parent <- rt2$go == rt2$parent
  rt2$size[is_parent & (rt2$size == 0 | is.na(rt2$size))] <-
    cluster_total[as.character(rt2$cluster[is_parent & (rt2$size == 0 | is.na(rt2$size))])]
}

# Plot with smaller labels (more will fit)
treemapPlot(rt2, fontsize.labels = 8)
```


```{r}
rt2 <- reducedTerms

# for each cluster, compute total size of all terms in that cluster
cluster_total <- tapply(rt2$size, rt2$cluster, sum, na.rm = TRUE)

# assign that total size to the parent row (go == parent)
is_parent <- rt2$go == rt2$parent
rt2$size[is_parent] <- cluster_total[ as.character(rt2$cluster[is_parent]) ]

treemapPlot(rt2)

```



```{r}
reducedTerms[reducedTerms$go == "GO:0031099", c("term","size","score")]

rt2 <- reducedTerms

rt2$size[rt2$go == "GO:0031099"] <- 
  max(rt2$size) * 0.6   # inflate visually

treemapPlot(rt2)

```
```{r}
blocks <- reducedTerms[reducedTerms$go == reducedTerms$parent, ]
treemapPlot(blocks)

```



```{r}
pdf("treemap_reducedTerms_epithelial_up.pdf",
    width = 8,
    height = 6)

treemapPlot(reducedTerms)

dev.off()


```


```{r}
go_ids_down <- cluster_summary_down_bg$ID

simMatrix_down <- calculateSimMatrix(
  go_ids_down,
  orgdb = "org.Dm.eg.db", 
  ont   = "BP",
  method = "Rel"
)
scores_down <- setNames(-log10(cluster_summary_down_bg$qvalue), cluster_summary_down_bg$ID)

reducedTerms_down <- reduceSimMatrix(simMatrix_down,
                                scores_down,
                                threshold=0.7,
                                orgdb="org.Dm.eg.db")

treemapPlot(reducedTerms_down)

```

```{r}
pdf("treemap_reducedTerms_epithelial_down.pdf",
    width = 8,
    height = 6)

treemapPlot(reducedTerms_down)

dev.off()

```

